---
title: "Calculating Distance and Elo Ranking"
author: "Aoi Ogawa and Lizette Carpenter"
date: "2/28/2020"
output:
  html_document: 
      toc: true
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# importing libraries
#library(datasets)
library(maps)
library(tidyr)
library(tidyverse)
library(readr)
library(geosphere)
library(sp)
library(leaflet) 
library(magrittr)
library(janitor)
library(ggridges)
library(dplyr)
library(lubridate)
```

```{r}
library(datasets)
library(scales)
library(infer)
```

```{r}
set.seed(1234)
```


#Importing Data 
```{r, include=FALSE}
# importing data
#games<-read.csv("data/games_Smith.csv")
#pbp<-read.csv("data/pbp_Smith.csv")
#stadium_data_coordinates <- read.csv("stadium_data_coordinates.csv")
#visit_team_data <- read.csv("visit_team_data.csv")
#ns_data_coordinates <- read.csv("Ns_data_coordinates.csv")
#all_data_coordinates <- read.csv("All_data_coordinates.csv")
```

## games
```{r, include=FALSE}
# importing data
games <-read.csv("data/games_Smith.csv") %>% 
  clean_names() %>%
  mutate(game_date = str_replace(game_date, "T00:00:00Z", ""),
         game_date = as.Date(game_date, "%Y-%m-%d")) 
```

## pbp
```{r}
pbp<-read.csv("data/pbp_Smith.csv") %>% 
  clean_names() %>% 
  mutate(
    game_clock = as.numeric(as.period(ms(game_clock), unit = "sec")),
    game_clock = 
      case_when(quarter == 1 ~ game_clock,
                quarter == 2 ~ game_clock + 900,
                quarter == 3 ~ game_clock + 1800,
                quarter == 4 ~ game_clock + 2700,
                quarter == 5 ~ game_clock + 3600))
```

## data (games + pbp) 
```{r}
data <- games %>% 
  left_join(pbp, by="game_key")
#note: NAs for 2007-2010 in pbp
```

## stadium_data_coordinates 
```{r}
#cleaning df column names to have a common variable with games (home_team)
stadium_data_coordinates <- read.csv("stadium_data_coordinates.csv") %>% 
  clean_names() 
```

## visit_team_data 
```{r}
visit_team_data <- read.csv("visit_team_data.csv") 
``` 

## games_w_coordinates (games + stadium_data_coordinates) 
```{r}
# games_w_coordinates = games + stadium_data_coordinates 
games_w_coordinates <- games %>%
  inner_join(stadium_data_coordinates)
```

# games_w_coordinates (games + stadium_data_coordinates + visit_team_data)   
```{r}
# games_w_coordinates = games + stadium_data_coordinates + visit_team_data  
games_w_coordinates <- visit_team_data %>%
  clean_names() %>% 
  inner_join(games_w_coordinates)  

# clean games_w_coordinates 
games_w_coordinates <- games_w_coordinates %>%
  mutate(num = NULL) %>%
  rename(home_lat = longitude,
         home_long = latitude) 
```
```{r}
stats <- pbp %>% 
  summarise(min = min(game_key),
            max = max(game_key))
```

## fulldata_with_thurs (cleaning classes of data w/ is_thurs) 
```{r}
fulldata_with_thurs <- data %>%
  filter(game_key >= 54816 & game_key <= 58155) %>% 
  mutate(is_thurs = ifelse(game_day == "Thursday", 1, 0)) %>% 
  mutate(is_thurs = as.factor(is_thurs),
         stadium = as.character(stadium),
         home_team = as.character(home_team),
         visit_team = as.character(visit_team),
         location = as.character(location),
         season = as.factor(season),
         week = as.factor(week),
         game_weather = as.character(game_weather),
         wind_speed = as.character(wind_speed),
         wind_direction = as.character(wind_direction),
         quarter = as.factor(quarter),
         play_description = as.character(play_description),
         down = as.factor(down),
         is_scoring_play = as.factor(is_scoring_play)) 
```

## nfl_elo
```{r}
nfl_elo <- read.csv("nfl_elo.csv") %>% 
  clean_names() %>% 
  rename(
    home_club_code = team1, 
    visitor_club_code = team2) %>% 
  mutate(
    game_date = as.Date(date, "%Y-%m-%d"), 
    home_club_code = 
      case_when(
        as.character(home_club_code) == "HOU" ~ "HST",
        as.character(home_club_code) == "LAR" ~ "SL",
        as.character(home_club_code) == "WSH" ~ "WAS", 
        as.character(home_club_code) == "CLE" ~ "CLV",
        as.character(home_club_code) == "BAL" ~ "BLT",
        as.character(home_club_code) == "ARI" ~ "ARZ",
        as.character(home_club_code) == "LAC" & as.character(season) >= "2017" ~ "SD",
        as.character(home_club_code) == "LAC" & as.character(season) <= "2016" ~ "LA",
        TRUE ~ as.character(home_club_code)), 
    visitor_club_code = 
      case_when(
        as.character(visitor_club_code) == "HOU" ~ "HST",
        as.character(visitor_club_code) == "LAR" ~ "SL",
        as.character(visitor_club_code) == "WSH" ~ "WAS", 
        as.character(visitor_club_code) == "CLE" ~ "CLV",
        as.character(visitor_club_code) == "BAL" ~ "BLT",
        as.character(visitor_club_code) == "ARI" ~ "ARZ",
        as.character(visitor_club_code) == "LAC" & as.character(season) >= "2017" ~ "SD",
        as.character(visitor_club_code) == "LAC" & as.character(season) <= "2016" ~ "LA",
        TRUE ~ as.character(visitor_club_code)))
```

```{r}
for_elo <- fulldata_with_thurs %>% 
  select(
    game_key,
    season, 
    game_number, 
    week, 
    game_date, 
    game_day, 
    home_club_code, 
    visitor_club_code, 
    ep, 
    is_thurs) 
```

## full_with_elo 
Combines NFL data (games and pbp) with FiveThirtyEight Elo data
```{r}
#full_with_elo 
full_with_elo= merge(nfl_elo, 
                     fulldata_with_thurs, 
                     by.x= c("home_club_code", "visitor_club_code", "game_date"), 
                     by.y= c("home_club_code", "visitor_club_code", "game_date"))

full_with_elo <- full_with_elo %>% 
  mutate(
    season = season.x,
    season.x = NULL, 
    season.y = NULL,
    is_home = ifelse(as.character(home_club_code) == as.character(possession_team), 1, 0),
    is_visitor =
      case_when(is_home == 1 ~ 0,
                is_home == 0 ~ 1),
    ep1 = 
      case_when(
        is_home == 1 ~ ep, 
        is_home == 0 ~ 0),
    ep2 = 
      case_when(
        is_visitor == 1 ~ ep,
        is_visitor == 0 ~ 0)) %>% 
  distinct(game_key, .keep_all = TRUE)
# new variables ep1 and ep2: expected points for home team (ep1) and expected points for visitor team (ep2).
# The expected points for only the possesion team changes play by play. The expected points for the defensive team would be 0.  
# ep1 and ep2 lets us see the expected points for the possession team for each play in a game. 
```

```{r}
# create domestic games df 
domestic_games <- games_w_coordinates %>%
  filter(
    stadium != c(
      'Wembley Stadium',
      "Rogers Centre",
      "Wembley",
      "Twickenham",
      "Estadio Azteca",
      "Twickenham Stadium",
      "Azteca Stadium",
      "Tottenham Hotspur Stadium",
      "Tottenham Hotspur"
    )
  )
# create international games df 
#international_games <- games_w_coordinates %>% 
#  filter(
#    stadium == c(
#      'Wembley Stadium',
#      "Rogers Centre",
#      "Wembley",
#      "Twickenham",
#      "Estadio Azteca",
#      "Twickenham Stadium",
#      "Azteca Stadium",
#      "Tottenham Hotspur Stadium",
#      "Tottenham Hotspur"
#    )
#  )
```

###Calculating and Categorizing Distances Travlled for Away Games
```{r}
# calculate distance
calc_domestic <- domestic_games %>% 
  rowwise() %>%
  mutate(distance_m = distCosine(c(home_long, home_lat),c(visit_long, visit_lat)),
         distance_mile = distance_m*0.6214/1000)

# categorize distances 
calc_domestic <- calc_domestic %>%
  mutate(
    distance =
      case_when(
        distance_mile > 2000 ~ "very far",
        distance_mile > 1000 ~ "far",
        distance_mile <= 1000 ~ "close"
      ),
    color_dis =
      case_when(
        distance == "very far" ~ "red",
        distance == "far" ~ "orange",
        distance == "close" ~ "cyan"
      ),
    is_thurs =
      ifelse(game_day == "Thursday", 1, 0),
    stroke =
      ifelse(is_thurs == '1', "8", "NULL"),
    home_team = as.character(home_team),
    visit_team = as.character(visit_team))
```

```{r}
#WITHOUT ELO RANKINGS
mapping_travel <- function(Season, Team) {
  mapping_team <- calc_domestic %>%
    filter(season == Season,
           visit_team == Team) 
   
  travel_map <- mapping_team %>%
    leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    setView(-93.65, 42.0285, zoom = 4.3) %>%
    addMarkers( ~ visit_long, ~ visit_lat) %>%
    addMarkers( ~ home_long, ~ home_lat) %>%
    addMarkers(
      lng = mapping_team$visit_long,
      lat = mapping_team$visit_lat,
      label = mapping_team$visit_team,
      labelOptions = labelOptions(noHide = T))
  
  for (i in 1:nrow(mapping_team)) {
    travel_map <- travel_map %>%
      addPolylines(
        lat = c(mapping_team[i, ]$visit_lat,
                mapping_team[i, ]$home_lat),
        lng = c(mapping_team[i, ]$visit_long,
                mapping_team[i, ]$home_long),
        color = mapping_team[i, ]$color_dis,
        dashArray = mapping_team[i,]$stroke
      ) %>%
      addMarkers(
        lng = mapping_team[i, ]$home_long,
        lat = mapping_team[i, ]$home_lat,
        label = mapping_team[i, ]$home_team,
        labelOptions = labelOptions(noHide = T)
      )
      # addMarkers(
      #   lng = mapping_team[i, ]$home_long,
      #   lat = mapping_team[i, ]$home_lat,
      #   popup = paste(
      #     mapping_team[i, ]$home_team,
      #     "<br>",
      #     mapping_team[i, ]$visit_team,
      #     "Ranking:",
      #     #round(mapping_team[i, ]$elo2_pre, digits = 0),
      #     #"<br>",
      #     #mapping_team[i, ]$home_team,
      #     #"Ranking:",
      #     #round(mapping_team[i, ]$elo1_pre, digits = 0),
      #     #"<br>"
      #   ),
      #   options = popupOptions(closeButton = TRUE)
      # )
  }
  
  return(travel_map)
  
}
mapping_travel("2016", "Los Angeles Rams")
```

### Adding Elo Rankings to Map 
```{r}
full_with_distances <- inner_join(full_with_elo, calc_domestic, by = c("home_team", "visit_team", "game_date"), copy = FALSE)

#selecting necessary varaibles for calculations and rankings
full_with_distances <- full_with_distances %>%
  rename(
    game_key = game_key.x, 
    season_type = season_type.x, 
    game_number = game_number.x,
    week = week.x,
    game_day = game_day.x, 
    game_site = game_site.x, 
    start_time = start_time.x, 
    stadium = stadium.x,
    location = location.x, 
    turf = turf.x, 
    time_zone = time_zone.x, 
    game_length = game_length.x, 
    game_weather = game_weather.x, 
    temperature = temperature.x, 
    humidity = humidity.x,
    wind_speed = wind_speed.x,
    wind_direction = wind_direction.x, 
    wind_chill = wind_chill.x, 
    distance = distance.x,
    is_thurs = is_thurs.x, 
    season = season.x) %>%
  mutate(
    game_key.x = NULL, 
    game_key.y = NULL, 
    season_type.x = NULL, 
    season_type.y = NULL, 
    game_number.x = NULL,
    game_number.y = NULL, 
    week.x = NULL,
    week.y = NULL, 
    game_day.x = NULL,
    game_day.y = NULL, 
    game_site.x = NULL,
    game_site.y = NULL, 
    start_time.x = NULL, 
    start_time.y = NULL,
    stadium_type.x = NULL,
    stadium_type.y = NULL, 
    location.x = NULL, 
    location.y = NULL, 
    stadium.x = NULL, 
    stadium.y = NULL,
    turf.x = NULL, 
    turf.y = NULL, 
    time_zone.x = NULL, 
    time_zone.y = NULL, 
    game_length.x = NULL, 
    game_length.y = NULL, 
    game_weather.x = NULL, 
    game_weather.y = NULL,
    temperature.x = NULL, 
    temperature.y = NULL, 
    humidity.x = NULL,
    humidity.y = NULL,
    wind_speed.x = NULL,
    wind_speed.y = NULL,
    wind_direction.x = NULL, 
    wind_direction.y = NULL, 
    wind_chill.x = NULL, 
    wind_chill.y = NULL,
    distance.x = NULL,
    distance.y = NULL,
    is_thurs.x = NULL, 
    is_thurs.y = NULL,
    season.x = NULL, 
    season.y = NULL)
```

```{r}
for_ranking <- full_with_distances %>%
  select(
    game_key,
    season,
    visit_long,
    visit_lat,
    home_long,
    home_lat,
    home_club_code, 
    visitor_club_code, 
    home_team,
    visit_team,
    color_dis,
    elo1_pre,
    elo2_pre,
    elo_prob1, 
    elo_prob2,
    is_thurs)
```

```{r}
mapping_travel <- function(Season, Team) {
  #obtain all of visiting team's travel games in a given season 
  mapping_team <- full_with_distances %>%
    filter(season == Season,
           visit_team == Team)
  
  #create leaflet map 
  travel_map <- mapping_team %>%
    leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    setView(-93.65, 42.0285, zoom = 4.3) %>%
    addMarkers( ~ visit_long, ~ visit_lat) %>%
    addMarkers( ~ home_long, ~ home_lat) %>%
    addMarkers(
      lng = mapping_team$visit_long,
      lat = mapping_team$visit_lat,
      label = mapping_team$visit_team,
      labelOptions = labelOptions(noHide = T)
    )
  
  for (i in 1:nrow(mapping_team)) {
    travel_map <- travel_map %>%
      addPolylines(
        lat = c(mapping_team[i, ]$visit_lat,
                mapping_team[i, ]$home_lat),
        lng = c(mapping_team[i, ]$visit_long,
                mapping_team[i, ]$home_long),
        color = mapping_team[i, ]$color_dis,
        dashArray = mapping_team[i,]$stroke
      ) %>%
      addTiles() %>%
      addMarkers(
        lng = mapping_team[i, ]$home_long,
        lat = mapping_team[i, ]$home_lat,
        popup = paste(
          mapping_team[i, ]$home_team,
          "<br>",
          mapping_team[i, ]$visit_team,
          "Ranking:",
          round(mapping_team[i, ]$elo2_pre, digits = 0),
          "<br>",
          mapping_team[i, ]$home_team,
          "Ranking:",
          round(mapping_team[i, ]$elo1_pre, digits = 0),
          "<br>"
        ),
        options = popupOptions(closeButton = TRUE)
      )
    travel_map
  }
  return(travel_map)
}

mapping_travel("2019", "Seattle Seahawks")
```

### Winning Probability Function 

Pr(A) = 1 / 10 * (-Elo Difference / 400) + 1

elo_prob1 | Home team's probability of winning according to Elo ratings
elo_prob2 | Away team's probability of winning according to Elo ratings

```{r}
win_prob <- function(Season, Team) { 
  
  #obtain data
  elo_rankings <- full_with_distances %>%
    filter(season == Season, 
           visit_team == Team) %>% 
    mutate(elo_diff = abs(elo1_pre - elo2_pre), 
           home_wp = 1/ (10 * (elo_diff/400) + 1)) %>% 
    select(season,
           home_team,
           visit_team,
           elo1_pre,
           elo2_pre,
           home_wp,
           elo_prob1,
           elo_prob2)
  
  return(elo_rankings)
}

win_prob("2019", "Seattle Seahawks")
```

```{r}
# function to get sem
se <- function(x) sqrt(var(x)/length(x))
```

```{r}
elo_visitors <- for_ranking %>% 
  filter(season >= 2007) %>% 
  group_by(visitor_club_code, is_thurs) %>% 
  summarise(n = n(),
            sum = sum(elo2_pre), 
            mean = sum / n, 
            avg = mean(elo2_pre),
            sem = se(elo2_pre))  %>% 
  mutate(is_thurs = 
           recode(is_thurs, 
                  "1" = "Thursday games",
                  "0" = "Not Thursday games"),
         is_thurs = as.character(is_thurs))

elo_visitors
```

```{r}
elo_home <- for_ranking %>% 
  filter(season >= 2007) %>% 
  group_by(home_club_code, is_thurs) %>% 
  summarise(n = n(),
            sum = sum(elo1_pre), 
            mean = sum / n, 
            avg = mean(elo1_pre),
            sem = se(elo1_pre)) %>% 
    mutate(is_thurs = 
           recode(is_thurs, 
                  "1" = "Thursday games",
                  "0" = "Not Thursday games"),
         is_thurs = as.character(is_thurs))
elo_home
```

```{r}
#permutation test (HOME TEAMS)  
#mean difference thurs - not thurs 
diff_elo_home <- elo_home %>% 
  specify(avg ~ is_thurs) %>% 
  calculate("diff in means", order = c("Not Thursday games", "Thursday games"))
diff_elo_home

#generate samples 
null_elo_home <- elo_home %>% 
  specify(avg ~ is_thurs) %>% 
  hypothesize(null= "independence") %>% 
  generate(reps = 5000, type = "permute") %>% 
  calculate("diff in means", order = c("Not Thursday games", "Thursday games"))
null_elo_home

null_elo_home %>% 
  get_pvalue(obs_stat = diff_elo_home, direction = "both")
null_elo_home

# the p-value is 0.926, which means that there is a 9.26% chance of seeing a difference at least as large as 1.64 in a world where there is no difference 
```

```{r}
#permutation test (VISITOR TEAMS)  
#mean difference thurs - not thurs 
diff_elo_visitor <- elo_visitors %>% 
  specify(avg ~ is_thurs) %>% 
  calculate("diff in means", order = c("Not Thursday games", "Thursday games"))
diff_elo_visitor

#generate samples 
null_elo_visitors <- elo_visitors %>% 
  specify(avg ~ is_thurs) %>% 
  hypothesize(null= "independence") %>% 
  generate(reps = 5000, type = "permute") %>% 
  calculate("diff in means", order = c("Not Thursday games", "Thursday games"))
null_elo_visitors

null_elo_visitors %>% 
  get_pvalue(obs_stat = diff_elo_home, direction = "both")
null_elo_visitors

# the p-value is 0.926, which means that there is a 9.26% chance of seeing a difference at least as large as 1.64 in a world where there is no difference 
```


```{r,message=FALSE}
ggplot(elo_home, aes(x = avg, y = is_thurs, fill = is_thurs)) +
  stat_density_ridges(quantile_lines = TRUE,
                      quantiles = 2, 
                      scale = 3, 
                      color = "white") + 
  scale_fill_manual(values = c("#E69F00", "#56B4E9"),
                    guide = FALSE) +
  labs(x = "Average Elo Ranking For Home Teams (2007 to 2019)", y = NULL) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

```{r}
null_elo_home %>% 
  visualise() +
  shade_p_value(obs_stat = diff_elo_home ,direction = "both") + 
  labs(x = "Difference in Mean Elo For Home Teams \n(Not Thursday games − Thursday games)",
       y = "Count",
       subtitle = "Red line shows observed difference in elo ranking") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank()) 
```

```{r,message=FALSE}
ggplot(elo_visitors, aes(x = avg, y = is_thurs, fill = is_thurs)) +
  stat_density_ridges(quantile_lines = TRUE, quantiles = 2, scale = 3, color = "white") + 
  scale_fill_manual(values = c("#E69F00", "#56B4E9"), guide = FALSE) +
  labs(x = "Average Elo Ranking For Visitor Teams (2007 to 2019)", y = NULL) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

```{r}
null_elo_visitors %>% 
  visualise() +
  shade_p_value(obs_stat = diff_elo_visitor ,direction = "both") + 
  labs(x = "Difference in Mean Elo For Visitor Teams \n(Not Thursday games − Thursday games)",
       y = "Count",
       subtitle = "Red line shows observed difference in elo ranking") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank()) 
```


```{r}
# all_elo_avg_diff <- for_ranking %>% 
#   filter(season >= 2007) %>% 
#   mutate(diff = abs(elo1_pre - elo2_pre)) %>% 
#   group_by(is_thurs) %>% 
#   summarise(n = n(), 
#             avg = mean(diff))
# all_elo_avg_diff      
# 
# ggplot(all_elo_avg_diff, aes(x = avg, y = is_thurs, fill = is_thurs)) +
#   stat_density_ridges(quantile_lines = TRUE, quantiles = 2, scale = 3, color = "white") + 
#   scale_fill_manual(values = c("#E69F00", "#56B4E9"), guide = FALSE) +
#   labs(x = "Average difference in elo ranking between teams in games from 2007 to 2019", y = NULL) +
#   theme_minimal() +
#   theme(panel.grid.minor = element_blank())
```

```{r}
elo_visitor_distance <- for_ranking %>% 
  filter(season >= 2007) %>% 
  group_by(visitor_club_code, color_dis) %>% 
  summarise(n = n(),
            sum = sum(elo2_pre), 
            mean = sum / n, 
            avg = mean(elo2_pre)) %>% 
  mutate(color_dis = 
           recode(color_dis, 
                  "cyan" = "close \n less than 1000 mi",
                  "orange" = "far \n less than 2000 mi ",
                  "red" =  "very far \n greater than 2000 mi"))
elo_visitor_distance 
```

```{r,message=FALSE}
ggplot(elo_visitor_distance, aes(x = avg, y = color_dis, fill = color_dis)) +
  stat_density_ridges(quantile_lines = TRUE, quantiles = 2, scale = 3, color = "white") + 
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#FF95FF"), guide = FALSE) +
  labs(x = "Average elo ranking from 2007 to 2019 \n For Visitor Teams", y = NULL) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

