---
title: "Distance and Accuracy"
author: "Elaona Lemoto"
date: "3/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(infer)
set.seed(9)
```

```{r}
# importing data
games<-read.csv("data/games_Smith.csv")
pbp<-read.csv("data/pbp_Smith.csv")
stadium_data_coordinates <- read.csv("stadium_data_coordinates.csv")
visit_team_data <- read.csv("visit_team_data.csv")
```

#Join DF's
```{r}
data<-games %>% 
  right_join(pbp, by="GameKey")
games_w_coordinates <-stadium_data_coordinates %>%
  inner_join(visit_team_data)
```

```{r}
#Create Null Observations
games_w_coordinates$num <-NULL
games_w_coordinates$additional_info_location <- NULL
colnames(games_w_coordinates)[colnames(games_w_coordinates) == 'longitude'] <- 'Home_lat'
colnames(games_w_coordinates)[colnames(games_w_coordinates) == 'latitude'] <- 'Home_long'
```

```{r}
#Counting different types of plays per game 

data$PlayDescription<-str_to_lower(data$PlayDescription)
data<-data %>% 
  filter(PlayNullifiedByPenalty == "N") %>% 
  mutate(pass_c= ifelse(PassResult == "C",1,0),
         pass=ifelse(str_detect(PlayDescription, "pass"), 1,0),
         run = ifelse(PlayResult>0 & (PassResult=="R" & is.na(SpecialTeamsPlayType)),1,0 ),
         bad_run = ifelse(PlayResult <= 0 & (PassResult == "NULL" | PassResult == "R") & is.na(SpecialTeamsPlayType), 1, 0 ))


data<-data %>% 
  group_by(PossessionTeam,GameKey, Season) %>% 
  summarize(total_cpass= sum(pass_c), 
            total_passes = sum(pass), 
            total_run = sum(run), 
            total_badrun = sum(bad_run)) %>% 
  mutate(completion_percentage = ((total_cpass)/total_passes), 
         pos_run_percent = (total_run/(total_run+total_badrun)))
```


```{r}
games_w_coordinates<-games_w_coordinates%>% 
  rowwise() %>%
  mutate(distance_km = distCosine(c(Home_long, Home_lat),c(Visit_long, Visit_lat)))%>%
  mutate(distance_mile = distance_km*0.6214/1000)

games_w_coordinates <-games_w_coordinates%>%
  mutate(distance = ifelse(distance_mile > 2000, "very far",
                           ifelse(distance_mile > 1000 && distance_mile <= 2000, "far", "close")))

prelim<-games %>% 
  left_join(games_w_coordinates, by = c("Home_Team", "Visit_Team", "Season", "Stadium"))

full_data<-prelim %>% 
  left_join(data, by = "GameKey") %>% 
  filter(Season.x>=2010)

full_data<-full_data %>% 
  mutate(is_thurs = ifelse(Game_Day == "Thursday", 1, 0))
thursday_games<-full_data %>% 
  filter(is_thurs==1)
nonthurs_games<-full_data %>% 
  filter(is_thurs==0)

```

#Exploratory Data Analysis
```{r}
attach(full_data)
par(mfrow=c(1,1))
summary(completion_percentage)
sd(completion_percentage)
summary(is_thurs)
sd(is_thurs)
summary(pos_run_percent)
sd(pos_run_percent)

```



```{r}
#Histograms

ggplot(full_data, aes(completion_percentage))+
  geom_histogram()

ggplot(full_data, aes(pos_run_percent))+
  geom_histogram()

ggplot(thursday_games, aes(completion_percentage))+
  geom_histogram()

ggplot(thursday_games, aes(pos_run_percent))+
  geom_histogram()

ggplot(nonthurs_games, aes(completion_percentage))+
  geom_histogram()

ggplot(nonthurs_games, aes(pos_run_percent))+
  geom_histogram()

ggplot(nonthurs_games, aes(distance_mile))+
  geom_histogram()

ggplot(thursday_games, aes(distance_mile))+
  geom_histogram()

ggplot(full_data, aes(distance_mile))+
  geom_histogram()


```

#Boxplots

```{r}

ggplot(full_data,aes(as.factor(is_thurs),completion_percentage))+
  geom_boxplot()+
  labs(title = "Pass Accuracy by Thursday and Non-Thursday games")

ggplot(full_data,aes(as.factor(is_thurs),pos_run_percent))+
  geom_boxplot()+
  labs(title = "Run Accuracy by Thursday and Non-Thursday games")

ggplot(full_data,aes(as.factor(distance),completion_percentage))+
  geom_boxplot()+
  labs(title = "Pass Accuracy by Distance")

ggplot(full_data,aes(as.factor(distance),pos_run_percent))+
  geom_boxplot()+
  labs(title = "Run Accuracy by Distance")
```


#Creating Df's for Permutation Tests for Thursday Games 
```{r}
# function to get sem
se <- function(x) (sd(x)/sqrt(length(x)))
#Cleaning Datasets 
# Create a dataset which summarised the total travel distance on Thursday or not Thursday and Distance

stats_team_season1 <- full_data%>%
  group_by(PossessionTeam, Season.x,is_thurs) %>%
  summarise(num_game=n(),
            sem_pass=se(completion_percentage),
            sem_run=se(pos_run_percent),
            avg_compl_perc = mean(completion_percentage), 
            avg_pos_run_perc = mean(pos_run_percent)) %>% 
  filter(Season.x>=2010) 

    
```

```{r}
# get the standard error mean for completion
all_games_sem_comp <- full_data %>%
  group_by(PossessionTeam,is_thurs) %>%
  summarise(sem = se(completion_percentage)) 

stats_team_season1$sem_pass[is.na(stats_team_season1$sem_pass)] <- 0

all_games_pos_run <- full_data %>%
  group_by(PossessionTeam,is_thurs) %>%
  summarise(sem = se(pos_run_percent)) 

stats_team_season1$sem_run[is.na(stats_team_season1$sem_run)] <- 0


```

#Permutations for Completion Percentage and Thursday Games 
```{r}
#Thursday and non-thursday
stats_team_season1<-stats_team_season1%>% 
  mutate(is_thurs = recode(is_thurs,
                         "1" = "Thursday games",
                         "0" = "Not Thursday games")) %>%
  mutate(is_thurs = as.character(is_thurs))


#Creating the Mean Diff Statistic
diff_coml_perc_thurs<-stats_team_season1 %>% 
  mutate(is_thurs = as.factor(is_thurs)) %>% 
  specify(avg_compl_perc ~ is_thurs) %>% 
  calculate("diff in means", order = c("Not Thursday games", "Thursday games"))

#Permuting over n-samples
permutations1<-stats_team_season1 %>%
  specify(avg_compl_perc ~ is_thurs) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 5000, type = "permute") %>%
  calculate("diff in means", order = c("Not Thursday games", "Thursday games"))

#Calculating P-value
permutations1 %>% 
  get_pvalue(obs_stat = diff_coml_perc_thurs, direction = "both")
```
We find that the associated p-value after running our permutation test is .256. Therefore we do not see statistically significant differences in pass accuracies between Thursday and non-Thursday games. Given our boxplot earlier, this does make sense. 

Now let's check for positive run percentages

#Permutations for Positive Run Percentages and Thursday Games
```{r}
#Creating the Mean Diff Statistic
diff_pos_run_perc<-stats_team_season %>% 
  mutate(is_thurs = as.factor(is_thurs)) %>% 
  specify(avg_pos_run_perc ~ is_thurs) %>% 
  calculate("diff in means", order = c("Not Thursday games", "Thursday games"))

#Permuting over n-samples
permutations2<-stats_team_season1 %>%
  specify(avg_pos_run_perc ~ is_thurs) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 5000, type = "permute") %>%
  calculate("diff in means", order = c("Not Thursday games", "Thursday games"))

#Calculating P-value
permutations2%>% 
  get_pvalue(obs_stat = diff_pos_run_perc, direction = "both")



```

Again, we see no statistical difference. However, we see that there is a lower probabiity of the average game existing within the nonthursday games average of positive runs than the p-value we found for completion percentage. 

#Creating Separate DF's for Distance and Permutation Test for Completion Percentage

```{r}
stats_team_season2 <- full_data%>%
  group_by(PossessionTeam, Season.x,distance) %>%
  summarise(num_game=n(),
            sem_pass=se(completion_percentage),
            sem_run=se(pos_run_percent),
            avg_compl_perc = mean(completion_percentage), 
            avg_pos_run_perc = mean(pos_run_percent)) %>% 
  filter(Season.x>=2010) 

#Creating the Mean Diff Statistic
diff_pos_run_perc<-stats_team_season2 %>% 
  mutate(distance = as.factor(distance)) %>% 
  specify(avg_pos_run_perc ~ distance) %>% 
  calculate("diff in means")

#Permuting over n-samples
permutations3<-stats_team_season1 %>%
  specify(avg_pos_run_perc ~ is_thurs) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 5000, type = "permute") %>%
  calculate("diff in means", order = c("Not Thursday games", "Thursday games"))

#Calculating P-value
permutations3%>% 
  get_pvalue(obs_stat = diff_pos_run_perc, direction = "both")


```


