---
title: "epa"
author: "Anna Ballou"
date: "5/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#load libraries
library(tidyverse)
library(janitor)
```

**Data:** 
```{r}
#read in and clean data.frames 
pbp <- read.csv("data/pbp_Smith.csv") %>% 
  clean_names()

games <-read.csv("data/games_Smith.csv") %>% 
  clean_names() %>%
  mutate(game_date = str_replace(game_date, "T00:00:00Z", ""),
         game_date = as.Date(game_date, "%Y-%m-%d")) 

data <- games %>% 
  inner_join(pbp, by="game_key")

data_thurs <- games %>% 
  inner_join(pbp, by="game_key") %>% 
  filter(game_day == "Thursday")
```

```{r}
#create and epa variable and subsequent data.frame with epa
epa_df <- data %>% 
  mutate(quarter = as.numeric(quarter)) %>% 
  mutate(half = ifelse(quarter < 3, 1, ifelse(quarter == 5, 3, 2))) %>%
  mutate(is_thurs = ifelse(game_day == "Thursday", 1, 0)) %>% 
  mutate(is_thurs = as.factor(is_thurs)) %>% 
  #grouping by game and half because EPA does not carry over halves or games
  group_by(game_key, half) %>%
  #make sure plays are in correct order
  arrange(game_key, play_id) %>%
  #calculating EPA - ADDED: if possession changes EPA must be calculated differently
  mutate(
    #actual points added
    points_added = ifelse(home_club_code == possession_team,
                         #if home team has the ball its the home points gained - visitor points gained
                         home_score_after_play - home_score_before_play - (visitor_score_after_play - visitor_score_before_play),
                         #otherwise its the visitor points gained - home points gained
                         visitor_score_after_play - visitor_score_before_play - (home_score_after_play - home_score_before_play)),
    #adding extra point to TDs - leaving rest of points added alone
    points_added = ifelse(points_added == 6, 7, points_added),
    points_added = ifelse(points_added == -6, -7, points_added),
    #calculating EPA - ADDED: if possession changes EPA must be calculated differently, removing Extra Point / kickoff plays
    epa = 
      case_when(#when play is a scrimmage play, no scoring and no change of possession
        down != 0 & points_added == 0 & possession_team == lead(possession_team) ~ lead(ep) - ep,
        #when a play is a scrimmage play, no scoring and a change of possession
        down != 0 & points_added == 0 & possession_team != lead(possession_team) ~ -lead(ep) - ep,
        #when a play is a scrimmage play with scoring
        down != 0 & points_added != 0 ~ points_added - ep)
  ) %>%
  #ungrouping dataframe
  ungroup() %>% 
  select(epa, is_thurs)
```

**EDA:**
```{r}
#view epa distribution
ggplot(data = epa_df, aes(x = epa)) +
  geom_density(kernel = "rectangular") +
  ggtitle("Distribution of EPA")

#plot epa on thurs vs sunday - density
ggplot(data = epa_df, aes(x = epa)) +
  geom_density(kernal = "rectangular", aes(color = is_thurs), size = 1.25) +
  ggtitle("Distribution of EPA", subtitle = "Comparing Thursday to Non Thursday Games") +
  coord_cartesian(xlim = c(-8, 8), ylim = c(0, 0.55)) +
  labs(color = "Thursday Status") +
  theme_minimal()

#plot epa on thurs vs sunday - histogram
ggplot(data = epa_df, aes(x = epa)) +
  geom_histogram(aes(color = is_thurs)) +
  ggtitle("Distribution of EPA by gameday", subtitle = "Histogram")

#plot epa on thurs vs sunday - boxplots
ggplot(data = epa_df, aes(y = epa, x = is_thurs)) +
  geom_boxplot(aes(color = is_thurs)) +
  ggtitle("comparing thursday and sunday games") +
  theme(legend.position = "none")
```

**Permutation test:** for difference in means (non-Thursday - Thursday)
```{r}
#set up
thurs <- epa_df %>% 
  filter(is_thurs == 1) %>% 
  select(is_thurs, epa)
dim(thurs)
nr <- epa_df %>% 
  filter(is_thurs == 0) %>% 
  select(is_thurs, epa)
dim(nr)
#cacluate original test stat: difference in mean EPA
mean_epa_thurs <- mean(thurs$epa, na.rm = T)
mean_epa_nr <- mean(nr$epa, na.rm = T)
(original_ts <-mean_epa_nr - mean_epa_thurs)
#regroup data
thurs_epa <- thurs %>% select(epa)
nr_epa <- nr %>% select(epa)
all_days <- rbind(thurs_epa, nr_epa)
all_days <- as.data.frame(all_days)
#permutations
distribution = NULL
set.seed(76)
for (i in 1:1000){
  temp = sample_frac(all_days, size = 1, replace = F)
  tempR = temp[1:26055,]
  tempNR = temp[26056:389014,]
  
  mean_R <- mean(tempR, na.rm = TRUE)
  mean_NR <- mean(tempNR, na.rm = TRUE)
  
  distribution[i] =  mean_NR - mean_R
  
}

sum(distribution >= original_ts)/length(distribution)
#p value = 0.935

#Plot of permutation test results
mean_differences <- as.data.frame(distribution) 
mean_val <- mean(distribution)
ggplot(data = mean_differences, aes(x = distribution)) + 
  geom_histogram(color = "blue", fill = "lightblue", bins = 15) +
  geom_vline(xintercept = original_ts, linetype = 2, color = "red") +
  ggtitle("Permutation test on EPA difference") +
  xlab("Non-Thursday EPA - Thursday EPA") +
  labs(caption = "red line = observed differene") +
  theme_minimal()
```

**Bootstrap Test:** non-Thursday - Thursday 
```{r}
#set up
bootstrap_epa <- epa_df %>% 
  select(is_thurs, epa) %>% 
  group_by(is_thurs) %>% 
  summarize(epa = mean(epa, na.rm = TRUE))
results = c()
#compute actual observed difference
observed_difference <- bootstrap_epa$epa[1] - bootstrap_epa$epa[2]
#set number of iterations and seed
num_iter = 500
set.seed(23)

#bootstrap
for(i in seq(1, num_iter)){
  #making a new df_total that will be randomly sample for the bootstrap
  df_total_temp <- epa_df %>% mutate(is_thurs = sample(is_thurs, size = nrow(epa_df), replace = T) )
  #splitting into thursday and not thursday and them summarizing results
  df_bootstrap <- df_total_temp %>% 
    group_by(is_thurs) %>% summarize(epa = mean(epa, na.rm = T))
  #appending data
  results = append(results, df_bootstrap$epa[2] - df_bootstrap$epa[1])
}
observed = data.frame(value = observed_difference, Line = "Observed Difference")

#plot
bootstrap_plot <- ggplot(as.data.frame(results), aes(results)) +
  #making histogram
  geom_histogram(colour = "blue", fill = "lightblue", bins = 10) +
  #adding line to represent observed data
  geom_vline(xintercept = observed_difference, linetype = 2, color = "red") +
  #setting xlabel
  xlab("Thursday EPA minus Non-Thursday EPA") + 
  #setting title
  ggtitle("Bootstrap Test on EPA Difference") +
  theme_minimal()
bootstrap_plot
```

**Examine EPA for teams in the AFC East & West:**
```{r}
#create team data.frame
epa_team_df <-  data %>% 
  filter(home_team == "Buffalo Bills" | home_team == "Miami Dolphins" | home_team == "New England Patriots" | home_team == "New York Jets" | home_team == "Kansas City Cheifs" | home_team == "Denver Broncos" | home_team == "Oakland Raiders") %>% 
  mutate(quarter = as.numeric(quarter)) %>% 
  mutate(half = ifelse(quarter < 3, 1, ifelse(quarter == 5, 3, 2))) %>%
  mutate(is_thurs = ifelse(game_day == "Thursday", 1, 0)) %>% 
  mutate(is_thurs = as.factor(is_thurs)) %>% 
  #grouping by game and half because EPA does not carry over halves or games
  group_by(game_key, half) %>%
  #make sure plays are in correct order
  arrange(game_key, play_id) %>%
  #calculating EPA - ADDED: if possession changes EPA must be calculated differently
  mutate(
    #actual points added
    points_added = ifelse(home_club_code == possession_team,
                          #if home team has the ball its the home points gained - visitor points gained
                          home_score_after_play - home_score_before_play - (visitor_score_after_play - visitor_score_before_play),
                          #otherwise its the visitor points gained - home points gained
                          visitor_score_after_play - visitor_score_before_play - (home_score_after_play - home_score_before_play)),
    #adding extra point to TDs - leaving rest of points added alone
    points_added = ifelse(points_added == 6, 7, points_added),
    points_added = ifelse(points_added == -6, -7, points_added),
    #calculating EPA - ADDED: if possession changes EPA must be calculated differently, removing Extra Point / kickoff plays
    epa = 
      case_when(#when play is a scrimmage play, no scoring and no change of possession
        down != 0 & points_added == 0 & possession_team == lead(possession_team) ~ lead(ep) - ep,
        #when a play is a scrimmage play, no scoring and a change of possession
        down != 0 & points_added == 0 & possession_team != lead(possession_team) ~ -lead(ep) - ep,
        #when a play is a scrimmage play with scoring
        down != 0 & points_added != 0 ~ points_added - ep)
  ) %>%
  #ungrouping dataframe
  ungroup() %>% 
  group_by(home_team, is_thurs, season) %>% 
  summarise(mean_epa = mean(epa, na.rm = T))
#plot
team_plot <- ggplot(data = epa_team_df, aes(x = home_team, y = mean_epa, fill = is_thurs)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle=90)) + 
  geom_hline(yintercept = 0, color = "black", size = 1) +
  ggtitle(label = "Mean EPA in the AFC East & West", subtitle = " Comparing  Thursday  vs. Non Thursday games") +
  labs(fill = "Thursday Status", y = "Mean EPA", x = "Home Team")
team_plot
```

**Examining EPA for teams with both winning and losing seasons:**
```{r}
#find which teams had winning and losing seasons
season_data <- data %>% 
  select(season, game_key, home_team, visit_team, home_score_after_play, visitor_score_after_play) %>% 
  group_by(game_key, home_team, visit_team, season) %>% 
  summarise(compare = case_when(last(home_score_after_play > visitor_score_after_play) ~ 1,
                             last(home_score_after_play == visitor_score_after_play) ~ 0,
                              TRUE ~ -1)) %>% 
  mutate(who_won = case_when(compare == 1 ~ "home", compare == 0 ~ "tie", TRUE ~ "visitor")) %>% 
  mutate(home_win = ifelse(who_won == "home", 1, 0),
         visit_win = ifelse(who_won == "visitor", 1, 0))

home_wins_data <- season_data %>% 
  group_by(home_team, season) %>% 
  summarize(num_home_wins = sum(home_win)) %>% 
  mutate(team = home_team)
  
visit_wins_data <- season_data %>% 
  group_by(visit_team, season) %>% 
  summarize(num_visit_wins = sum(visit_win)) %>% 
  mutate(team = visit_team)

season_wins <- home_wins_data %>% 
  left_join(visit_wins_data, by = c("team","season")) %>% 
  mutate(num_wins = num_home_wins + num_visit_wins,
         is_winning_season = ifelse(num_wins > 8, 1, 0))
head(season_wins)
```

Function for calculating and plotting EPA on thursday and non-Thursday for any team:
```{r}
team_epa <- function(team) {
 team_epa <-  data %>% 
    filter(home_team == team) %>% 
    mutate(quarter = as.numeric(quarter)) %>% 
    mutate(half = ifelse(quarter < 3, 1, ifelse(quarter == 5, 3, 2))) %>%
    mutate(is_thurs = ifelse(game_day == "Thursday", 1, 0)) %>% 
    mutate(is_thurs = as.factor(is_thurs)) %>% 
    #grouping by game and half because EPA does not carry over halves or games
    group_by(game_key, half) %>%
    #make sure plays are in correct order
    arrange(game_key, play_id) %>%
    #calculating EPA - ADDED: if possession changes EPA must be calculated differently
    mutate(
      #actual points added
      points_added = ifelse(home_club_code == possession_team,
                            #if home team has the ball its the home points gained - visitor points gained
                            home_score_after_play - home_score_before_play - (visitor_score_after_play - visitor_score_before_play),
                            #otherwise its the visitor points gained - home points gained
                            visitor_score_after_play - visitor_score_before_play - (home_score_after_play - home_score_before_play)),
      #adding extra point to TDs - leaving rest of points added alone
      points_added = ifelse(points_added == 6, 7, points_added),
      points_added = ifelse(points_added == -6, -7, points_added),
      #calculating EPA - ADDED: if possession changes EPA must be calculated differently, removing Extra Point / kickoff plays
      epa = 
        case_when(#when play is a scrimmage play, no scoring and no change of possession
          down != 0 & points_added == 0 & possession_team == lead(possession_team) ~ lead(ep) - ep,
          #when a play is a scrimmage play, no scoring and a change of possession
          down != 0 & points_added == 0 & possession_team != lead(possession_team) ~ -lead(ep) - ep,
          #when a play is a scrimmage play with scoring
          down != 0 & points_added != 0 ~ points_added - ep)
    ) %>%
    #ungrouping dataframe
    ungroup() %>% 
    mutate(season = as.factor(season)) %>% 
    group_by(home_team, is_thurs, season) %>% 
    summarise(mean_epa = mean(epa, na.rm = T))
  
  #plot
  team_box <- ggplot(data = team_epa, aes(x = is_thurs, y = mean_epa, fill = is_thurs)) +
    geom_boxplot() +
    ggtitle("Comparing EPA on Thurday and Non-Thursday games", subtitle = team) +
    labs(x = "Thursday status", y = "mean EPA")
  
  return(team_box)
}
```

3 Examples: Miami Dolphins (first), Atlanta Falcons (second), New England Patriots (third)
```{r}
team_epa("Miami Dolphins")
team_epa("Atlanta Falcons")
team_epa("New England Patriots")
```

