---
title: "Final Scores"
author: "Lizette Carpenter"
date: "4/27/2020"
output:
  html_document: 
      toc: true
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
```


This RMarkdown includes a function that outputs the final score and margin of victory for all the games of a given team in a given season. 

1. Part 1 includes the data wrangling that creates the dataframe used in this function. 
1. Part 2 compares the scores in the newly created dataframe with the score in FiveThirtyEight Elo Ranking dataframe. 
1. Part 3 includes the functions that outputs final scores and margin of victory. 

# Loading Libraries and Data 

```{r}
library(janitor)
library(lubridate)
library(dplyr)
library(tidyverse)
```

Before calculating the final scores, the following data frames are cleaned: 

```{r}
#games
games <-read.csv("data/games_Smith.csv") %>% 
  clean_names() %>%
  mutate(game_date = str_replace(game_date, "T00:00:00Z", ""),
         game_date = as.Date(game_date, "%Y-%m-%d")) 
```

```{r}
#pbp
pbp<-read.csv("data/pbp_Smith.csv") %>% 
  clean_names() %>% 
  mutate(
    game_clock = as.numeric(as.period(ms(game_clock), unit = "sec")),
    game_clock = 
      case_when(quarter == 1 ~ game_clock,
                quarter == 2 ~ game_clock + 900,
                quarter == 3 ~ game_clock + 1800,
                quarter == 4 ~ game_clock + 2700,
                quarter == 5 ~ game_clock + 3600))
```

```{r}
#data (combined)
data <- games %>% 
  left_join(pbp, by="game_key")
```

```{r}
#fulldata
fulldata <- data %>% 
  mutate(is_thurs = ifelse(game_day == "Thursday", 1, 0)) %>% 
  filter(game_day == "Thursday" | game_day == "Sunday") %>% 
  mutate(is_thurs = as.factor(is_thurs),
         stadium = as.character(stadium),
         home_team = as.character(home_team),
         visit_team = as.character(visit_team),
         location = as.character(location),
         game_key = as.factor(game_key), 
         season = as.factor(season),
         week = as.factor(week),
         game_weather = as.character(game_weather),
         wind_speed = as.character(wind_speed),
         wind_direction = as.character(wind_direction),
         quarter = as.factor(quarter),
         play_description = as.character(play_description),
         down = as.factor(down),
         is_scoring_play = as.factor(is_scoring_play))
```

```{r}
nfl_elo <- read.csv("nfl_elo.csv") %>% 
  clean_names() %>% 
  rename(
    home_club_code = team1, 
    visitor_club_code = team2) %>% 
  mutate(
    game_date = as.Date(date, "%Y-%m-%d"), 
    date = NULL, 
    home_club_code = 
      case_when(
        as.character(home_club_code) == "HOU" ~ "HST",
        as.character(home_club_code) == "LAR" ~ "SL",
        as.character(home_club_code) == "WSH" ~ "WAS", 
        as.character(home_club_code) == "CLE" ~ "CLV",
        as.character(home_club_code) == "BAL" ~ "BLT",
        as.character(home_club_code) == "ARI" ~ "ARZ",
        as.character(home_club_code) == "LAC" & as.character(season) >= "2017" ~ "SD",
        as.character(home_club_code) == "LAC" & as.character(season) <= "2016" ~ "LA",
        TRUE ~ as.character(home_club_code)), 
    visitor_club_code = 
      case_when(
        as.character(visitor_club_code) == "HOU" ~ "HST",
        as.character(visitor_club_code) == "LAR" ~ "SL",
        as.character(visitor_club_code) == "WSH" ~ "WAS", 
        as.character(visitor_club_code) == "CLE" ~ "CLV",
        as.character(visitor_club_code) == "BAL" ~ "BLT",
        as.character(visitor_club_code) == "ARI" ~ "ARZ",
        as.character(visitor_club_code) == "LAC" & as.character(season) >= "2017" ~ "SD",
        as.character(visitor_club_code) == "LAC" & as.character(season) <= "2016" ~ "LA",
        TRUE ~ as.character(visitor_club_code)))
```

# Part 1 

```{r}
#full data with final scores 
data_with_scores <- fulldata %>% 
  group_by(game_key, 
           home_club_code, 
           visitor_club_code) %>% 
  #Mutates home/visitor_score_after_play. The values of these variables will only show the 
  #team's score if the play is a scoring play. Otherwise, the value will be 0.  
  mutate( 
    home_score_after_play = 
      case_when(
        is_scoring_play == 0 ~ 0,
        TRUE ~ as.double(home_score_after_play)),
    visitor_score_after_play = 
      case_when(
        is_scoring_play == 0 ~ 0, 
        TRUE ~ as.double(visitor_score_after_play))) %>% 
  group_by(game_clock) %>% 
  #Filters out home/visitor_score_after_play with values that have 0.
  filter(!home_score_after_play == 0,
         !visitor_score_after_play == 0) %>%
  group_by(game_key) %>%
  mutate(home_final = max(home_score_after_play),
         visitor_final = max(visitor_score_after_play),
         winning_team = 
           case_when(
             home_final > visitor_final ~ as.character(home_club_code), 
             home_final < visitor_final ~ as.character(visitor_club_code),
             home_final == visitor_final ~ "tie"), 
         is_tie = ifelse(winning_team == "tie", 1, 0))
```

# Part 2 

This section is left commented out because it is not necessary to run part 3. The results of this code tells us that FiveThirtyEight and NFL data have the same data for final scores. 
```{r}
# nfl_elo_2019 <- nfl_elo %>% 
#   filter(season == "2019") 
# 
# nfl_fs <- data_with_scores %>% 
#   select(game_key, game_date, season, home_club_code, visitor_club_code, is_scoring_play, home_score_after_play, visitor_score_after_play, home_final, visitor_final, winning_team, is_tie) %>%
#   distinct(game_key, .keep_all = TRUE)
# 
# comparisons <- inner_join(nfl_fs, nfl_elo_2019, by = NULL, copy = FALSE) 
# #full_with_elo 
# comparisons= merge(nfl_elo_2019, 
#                    nfl_fs, 
#                    by.x= c("home_club_code", "visitor_club_code", "game_date", "season"), 
#                    by.y= c("home_club_code", "visitor_club_code", "game_date", "season"))
# 
# comparisons <- comparisons %>% 
#   select(game_date, season, home_club_code, visitor_club_code, home_final, visitor_final, score1, score2)
# 
# for (i in 1:nrow(comparisons)) {
#   home_team <- comparisons[i,]$home_club_code
#   vis_team <- comparisons[i,]$visitor_club_code
#   home_elo_score <- comparisons[i,]$score1
#   home_nfl_score <- comparisons[i,]$home_final
#   vis_elo_score <- comparisons[i,]$score2
#   vis_nfl_score <- comparisons[i,]$visitor_final
#   if (home_elo_score == home_nfl_score) {
#     home <- "yes"
#   } else {
#     home <- "no"
#   }
#   if (vis_elo_score == vis_nfl_score) {
#     vis <- "yes"
#   } else {
#     vis <- "no"
#   }
#   results <- c(home_team, vis_team, home, vis)
#   print(results)
# }
```

# Part 3 (Final Scores and Margin of Victory)

```{r}
final_score <- function(Season, Team) {
  record <- data_with_scores %>%
    filter(season == Season,
           (home_club_code == Team) |
             (visitor_club_code == Team)) %>%
    select(
      season,
      game_key,
      game_date,
      week, 
      game_day,
      game_clock,
      home_club_code,
      visitor_club_code,
      is_scoring_play,
      home_score_after_play,
      visitor_score_after_play,
      home_final,
      visitor_final,
      winning_team,
      is_tie
    ) %>%
    mutate(
      mv = abs(home_final - visitor_final)
    ) %>%
    select(
      season, 
      game_key,
      game_date,
      week,
      game_day, 
      home_club_code,
      visitor_club_code,
      home_final,
      visitor_final,
      mv
    ) %>% 
    distinct(game_date, .keep_all = TRUE) %>% 
    group_by(week) 
  return(record)
}
final_score("2019", "SEA")
```

# Season and Game EDA (in progress)

```{r}
summary <- function(Season, Team) {
  record <- data_with_scores %>%
    filter(season == Season,
           (home_club_code == Team) |
             (visitor_club_code == Team)) %>%
    select(
      season,
      week, 
      game_key,
      game_date,
      game_clock,
      home_club_code,
      visitor_club_code,
      is_scoring_play,
      home_score_after_play,
      visitor_score_after_play,
      home_final,
      visitor_final,
      winning_team,
      is_tie
    ) %>%
    mutate(
      mv = abs(home_final - visitor_final)
    ) %>%
    group_by(game_key, week, home_club_code, visitor_club_code) %>% 
    summarise(
      is_scoring_play = n()
      )
  return(record)
}
summary("2019", "SEA")
```

