---
title: "Travel distance vs margin of victory"
author: "Aoi Ogawa"
date: "5/7/2020"
output: html_document
---


## This document analyzed the relation between margin of victory and travel distance, and win/lose and travel distance


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### importing libraries
```{r, include = FALSE, warning=FALSE}
library(datasets)
library(maps)
library(tidyr)
library(tidyverse)
library(readr)
# install.packages("geosphere")
library(geosphere)
library(dplyr)
library(sp)
library(leaflet)
library(magrittr)
library(janitor)
library(lubridate)
#  install.packages("infer")
library(infer)
# install.packages("ggridges")
library(ggridges)
library(scales)

# setting seed
set.seed(1234)
```

### Importing data
```{r,warning=FALSE}
#game data
games <-read.csv("data/games_Smith.csv") 

#pbp data
pbp<-read.csv("data/pbp_Smith.csv") %>% 
  group_by(GameKey) %>%
  slice(n()) %>%
  ungroup 
  
short_pbp <- pbp %>%
  select(1,18,19,30)%>%
  mutate(margin_victory = HomeScoreAfterPlay - VisitorScoreAfterPlay)


# joining the two datasets  
win_mergin_data <- games %>%
  inner_join(short_pbp, by="GameKey") %>%
  mutate(is_thurs = ifelse(Game_Day == "Thursday","Thursday games","Not Thursday games"))

```


### Creating datasets
```{r, warning=FALSE}
#cleaning df column names to have a common variable with games (home_team)
# stadium coordinates data
stadium_data_coordinates <- read.csv("stadium_data_coordinates.csv") 


# visit team stadium coordinates data
visit_team_data <- read.csv("visit_team_data.csv")


# merging the datasets
mv_w_coordinates <- win_mergin_data %>%
  inner_join(stadium_data_coordinates )

mv_w_coordinates <- mv_w_coordinates %>%
  inner_join(visit_team_data)

mv_w_coordinates$num <-NULL
mv_w_coordinates$additional_info_location <- NULL
colnames(mv_w_coordinates)[colnames(mv_w_coordinates) == 'longitude'] <- 'Home_lat'
colnames(mv_w_coordinates)[colnames(mv_w_coordinates) == 'latitude'] <- 'Home_long'

mv_w_coordinates <- mv_w_coordinates[!duplicated(mv_w_coordinates$GameKey), ]

remove.list <- paste(c('Wembley Stadium',"Rogers Centre","Wembley","Twickenham","Estadio Azteca","Twickenham Stadium","Azteca Stadium","Tottenham Hotspur Stadium","Tottenham Hotspur"), collapse = '|')


# creating domestic games dataset
mv_domestic_games <- mv_w_coordinates %>%
  filter(!grepl(remove.list, Stadium)) %>%
  mutate(is_thurs = ifelse(Game_Day == "Thursday", 1, 0)) 

mv_domestic_games <- mv_domestic_games[!duplicated(mv_domestic_games$GameKey), ]

# creating international games dataset
mv_international_games <- mv_domestic_games %>%
  filter(grepl(remove.list, Stadium)) %>%
  mutate(is_thurs = ifelse(Game_Day == "Thursday", 1, 0)) 

mv_international_games <- mv_international_games[!duplicated(mv_international_games$GameKey), ]

```

### Calculating distance
```{r,warning=FALSE}
mv_calc_domestic <- mv_domestic_games %>% 
  rowwise() %>%
  mutate(distance_m = distCosine(c(Home_long, Home_lat),c(Visit_long, Visit_lat))) %>%
  mutate(distance_mile = distance_m*0.6214/1000)

mv_calc_domestic <- mv_calc_domestic[!duplicated(mv_calc_domestic$GameKey), ]

# separating distance into three groups (close, far, very far)
mv_calc_domestic <- mv_calc_domestic %>%
  mutate(distance = ifelse(distance_mile > 2000, "very far",
                           ifelse(distance_mile > 1000 && distance_mile <= 2000, "far", "close"))) 

```

### Creating datasets for 
```{r,warning=FALSE}
mv_sum_distance_all <- mv_calc_domestic %>%
  group_by(Visit_Team) %>%
  summarise(travel_dis =sum(distance_mile)) %>%
  rename(Team = Visit_Team)


mv_num_games_visit <- mv_calc_domestic %>%
  group_by(Visit_Team) %>%
  summarise(num_game_visit =n()) %>%
  rename(Team = Visit_Team)

mv_num_games_home <- mv_calc_domestic %>%
  group_by(Home_Team) %>%
  summarise(num_game_home =n()) %>%
  rename(Team = Home_Team)

mv_average_home <- mv_calc_domestic %>%
  group_by(Home_Team)%>%
  summarise(avg_home_mv = mean(margin_victory)) %>%
  rename(Team = Home_Team)

mv_average_visit <- mv_calc_domestic %>%
  group_by(Visit_Team)%>%
  summarise(avg_visit_mv = mean(margin_victory))%>%
  rename(Team = Visit_Team)


mv_distance_all <- mv_sum_distance_all %>%
  inner_join(mv_num_games_visit) %>%
  left_join(mv_num_games_home)%>%
  left_join(mv_average_home)%>%
  left_join(mv_average_visit)%>%
  mutate(total_game = num_game_visit + num_game_home) %>%
  mutate(avg_travel = travel_dis/total_game) %>%
  mutate(avg_mv = ((num_game_visit*avg_visit_mv)+(num_game_home*avg_home_mv))/total_game)

```

### Create a graph that shows the distribution of margin of error (home score - visit score)
```{r,warning=FALSE}
ggplot(win_mergin_data, aes(x = margin_victory)) +
  geom_density(color="darkblue",fill="lightblue") +
  geom_vline(aes(xintercept=mean(margin_victory)),
           color="blue", linetype="dashed", size=1) +
  labs(x="Margin of Victory (home score - visit score)", title = "The Distribution of Margin of Victory")

```


# Run a permutation test
```{r, warning=FALSE}
# test stat
mv_distance <- mv_distance_all %>% 
  specify(avg_mv ~ avg_travel) %>% 
  calculate(stat = "correlation")

lin_mod <- lm(avg_mv ~ avg_travel, data = mv_distance_all)
summary(lin_mod)

# generate samples 
null_mv_distance <- mv_distance_all %>%
  specify(avg_mv ~ avg_travel) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 5000, type = "permute") %>%
  calculate("correlation")

# run a permutation test
null_mv_distance %>% 
  get_pvalue(obs_stat = mv_distance, direction = "both")
# p-value = 0.442
```

### Generating plots
```{r,warning=FALSE}
# correlation between average travel distance and average margin of victory
ggplot(mv_distance_all, aes(x = avg_travel, y = avg_mv)) +
  geom_point() +
  geom_smooth(method=lm, se=FALSE) +
  labs(x = "Average Travel Distances from 2010 to 2019 (miles)", 
       y = "Average Margin of Victory \n (Home team score - visitor team score)", 
       title = "Correlation between Average Travel Distance and Average Margin of Victory")


# simulaiton based null distribution
null_mv_distance %>% 
  visualise() +
  shade_p_value(obs_stat = mv_distance,direction = "both") + 
  labs(x = "Correlation coefficient between\n average travel distance and average margin of victory",
       y = "Count",
       subtitle = "Red line shows observed correlaiton between\n average travel distance and average margin of victory") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank()) 
```

### Categorizing the travel distance 
```{r,warning=FALSE}
# filtered the data to just visit teams

mv_cat_distance_all <- mv_calc_domestic %>%
  group_by(Visit_Team, distance) %>%
  summarise(travel_dis =sum(distance_mile)) %>%
  rename(Team = Visit_Team)


mv_cat_num_games_visit <- mv_calc_domestic %>%
  group_by(Visit_Team, distance) %>%
  summarise(num_game_visit =n()) %>%
  rename(Team = Visit_Team)


mv_cat_average_visit <- mv_calc_domestic %>%
  group_by(Visit_Team, distance)%>%
  summarise(avg_visit_mv = mean(margin_victory))%>%
  rename(Team = Visit_Team)


mv_cat_distance_all <- mv_cat_distance_all %>%
  inner_join(mv_cat_num_games_visit) %>%
  left_join(mv_cat_average_visit) %>%
  mutate(avg_travel = travel_dis/num_game_visit) 
```

### run a permutation test 
```{r}
# obtaining the test stats
diff_cat_distance <- mv_cat_distance_all %>% 
  specify(avg_visit_mv ~ distance) %>% 
  calculate(stat = "F")

diff_cat_distance
# 0.258

# generate samples 
null_mv_travel <- mv_cat_distance_all %>%
  specify(avg_visit_mv ~ distance) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 5000, type = "permute") %>%
  calculate("F")

null_mv_travel

null_mv_travel %>% 
  get_pvalue(obs_stat = diff_cat_distance, direction = "right")

# the p-value is 0.774
```

### Generating plots
```{r,warning=FALSE}
# Plots that shows the distribution of average margin of victory
ggplot(mv_cat_distance_all, aes(x = avg_visit_mv, y = distance, fill = distance)) +
  stat_density_ridges(quantile_lines = TRUE, quantiles = 2, scale = 3, color = "white") + 
  scale_fill_manual(values = c("#E69F00", "#56B4E9","#999999"), guide = FALSE) +
  labs(x = "Average Margin of Victory from 2010 to 2019", y = NULL,
       title="Comparison of Average Margin of Victory for Visiting Teams \n (close < 1000 miles, 1000 <= far <= 2000, 2000 < very far)") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank()) 


# plots that shows the null distribution based on the simulation result
null_mv_travel %>% 
  visualise() +
  shade_p_value(obs_stat = diff_cat_distance, direction = "right") + 
  labs(x = "The F-value from comparing three different groups \n (close < 1000 miles, 1000 <= far <= 2000, 2000 < very far)",
       y = "Count",
       subtitle = "Red line shows observed variation of margin of victory") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank()) 
```


