---
title: "EP nd Rankings Visualizations"
author: "Lizette Carpenter"
date: "4/24/2020"
output:
  html_document: 
      toc: true
      collapsed: false
      smooth_scroll: false

---

# Load libraries 
```{r message = FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(ggridges)
```

# Data Wrangling: 

The following are dataframes needed to generate the visualizations. 

## games
```{r}
games <-read.csv("data/games_Smith.csv") %>% 
  clean_names() %>%
  mutate(game_date = str_replace(game_date, "T00:00:00Z", ""),
         game_date = as.Date(game_date, "%Y-%m-%d")) 
```

## pbp
```{r}
pbp<-read.csv("data/pbp_Smith.csv") %>% 
  clean_names() %>% 
  mutate(
    game_clock = as.numeric(as.period(ms(game_clock), unit = "sec")),
    game_clock = 
      case_when(quarter == 1 ~ game_clock,
                quarter == 2 ~ game_clock + 900,
                quarter == 3 ~ game_clock + 1800,
                quarter == 4 ~ game_clock + 2700,
                quarter == 5 ~ game_clock + 3600))
```

## data (combined) 
```{r}
data <- games %>% 
  left_join(pbp, by="game_key")
#note: NAs for 2007-2010 in pbp
```

## fulldata_with_thurs 
```{r}
fulldata_with_thurs <- data %>% 
  mutate(is_thurs = ifelse(game_day == "Thursday", 1, 0)) %>% 
  filter(game_day == "Thursday" | game_day == "Sunday") %>% 
  mutate(is_thurs = as.factor(is_thurs),
         stadium = as.character(stadium),
         home_team = as.character(home_team),
         visit_team = as.character(visit_team),
         location = as.character(location),
         game_key = as.factor(game_key), 
         season = as.factor(season),
         week = as.factor(week),
         game_weather = as.character(game_weather),
         wind_speed = as.character(wind_speed),
         wind_direction = as.character(wind_direction),
         quarter = as.factor(quarter),
         play_description = as.character(play_description),
         down = as.factor(down),
         is_scoring_play = as.factor(is_scoring_play))
```

## nfl_elo_2019
```{r}
nfl_elo_2019 <- read.csv("nfl_elo_latest.csv") %>% 
  clean_names() %>% 
  rename(
    home_club_code = team1, 
    visitor_club_code = team2) %>% 
  mutate(
    game_date = as.Date(date, "%Y-%m-%d"),
    home_club_code = 
      case_when(
        as.character(home_club_code) == "HOU" ~ "HST",
        as.character(home_club_code) == "LAR" ~ "SL",
        as.character(home_club_code) == "WSH" ~ "WAS", 
        as.character(home_club_code) == "CLE" ~ "CLV",
        as.character(home_club_code) == "BAL" ~ "BLT",
        as.character(home_club_code) == "ARI" ~ "ARZ",
        as.character(home_club_code) == "LAC" & as.character(season) >= "2017" ~ "SD",
        as.character(home_club_code) == "LAC" & as.character(season) <= "2016" ~ "LA",
        TRUE ~ as.character(home_club_code)), 
    visitor_club_code = 
      case_when(
        as.character(visitor_club_code) == "HOU" ~ "HST",
        as.character(visitor_club_code) == "LAR" ~ "SL",
        as.character(visitor_club_code) == "WSH" ~ "WAS", 
        as.character(visitor_club_code) == "CLE" ~ "CLV",
        as.character(visitor_club_code) == "BAL" ~ "BLT",
        as.character(visitor_club_code) == "ARI" ~ "ARZ",
        as.character(visitor_club_code) == "LAC" & as.character(season) >= "2017" ~ "SD",
        as.character(visitor_club_code) == "LAC" & as.character(season) <= "2016" ~ "LA",
        TRUE ~ as.character(visitor_club_code)))
```

## nfl_elo
```{r}
nfl_elo <- read.csv("nfl_elo.csv") %>% 
  clean_names() %>% 
  rename(
    home_club_code = team1, 
    visitor_club_code = team2) %>% 
  mutate(
    game_date = as.Date(date, "%Y-%m-%d"), 
    home_club_code = 
      case_when(
        as.character(home_club_code) == "HOU" ~ "HST",
        as.character(home_club_code) == "LAR" ~ "SL",
        as.character(home_club_code) == "WSH" ~ "WAS", 
        as.character(home_club_code) == "CLE" ~ "CLV",
        as.character(home_club_code) == "BAL" ~ "BLT",
        as.character(home_club_code) == "ARI" ~ "ARZ",
        as.character(home_club_code) == "LAC" & as.character(season) >= "2017" ~ "SD",
        as.character(home_club_code) == "LAC" & as.character(season) <= "2016" ~ "LA",
        TRUE ~ as.character(home_club_code)), 
    visitor_club_code = 
      case_when(
        as.character(visitor_club_code) == "HOU" ~ "HST",
        as.character(visitor_club_code) == "LAR" ~ "SL",
        as.character(visitor_club_code) == "WSH" ~ "WAS", 
        as.character(visitor_club_code) == "CLE" ~ "CLV",
        as.character(visitor_club_code) == "BAL" ~ "BLT",
        as.character(visitor_club_code) == "ARI" ~ "ARZ",
        as.character(visitor_club_code) == "LAC" & as.character(season) >= "2017" ~ "SD",
        as.character(visitor_club_code) == "LAC" & as.character(season) <= "2016" ~ "LA",
        TRUE ~ as.character(visitor_club_code)))
```

## full_with_elo 

Combines NFL data (games and pbp) with FiveThirtyEight Elo data
```{r}
for_elo <- fulldata_with_thurs %>% 
  select(game_key,
         season,
         season_type,
         game_clock,
         game_number,
         week,
         game_date,
         game_day,
         home_club_code,
         visitor_club_code,
         possession_team,
         ep,
         is_thurs)

#full_with_elo 
full_with_elo= merge(nfl_elo, 
                     for_elo, 
                     by.x= c("home_club_code", "visitor_club_code", "game_date"), 
                     by.y= c("home_club_code", "visitor_club_code", "game_date"))

full_with_elo <- full_with_elo %>% 
  mutate(
    season = season.x,
    season.y = NULL, 
    is_home = ifelse(as.character(home_club_code) == as.character(possession_team), 1, 0),
    is_visitor =
      case_when(is_home == 1 ~ 0,
                is_home == 0 ~ 1),
    ep1 = 
      case_when(
        is_home == 1 ~ ep, 
        is_home == 0 ~ 0),
    ep2 = 
      case_when(
        is_visitor == 1 ~ ep,
        is_visitor == 0 ~ 0))
# new variables ep1 and ep2: expected points for home team (ep1) and expected points for visitor team (ep2).
# The expected points for only the possesion team changes play by play. The expected points for the defensive team would be 0.  
# ep1 and ep2 lets us see the expected points for the possession team for each play in a game. 
```

# Filters 
## AFC Teams
```{r}
#filter only afc teams 
afc <- full_with_elo %>% 
  filter(home_club_code %in% c("BUF", "MIA", "NEP", "NYJ",
                             "BLT", "CIN", "CLE", "PIT",
                             "HOU", "IND", "JAC", "TEN",
                             "DEN", "KC", "OAK"),
         visitor_club_code %in% c("BUF", "MIA", "NEP", "NYJ",
                                "BLT", "CIN", "CLE", "PIT",
                                "HOU", "IND", "JAC", "TEN",
                                "DEN", "KC", "OAK"))
```

# Data Visualizations 

## Change in Expected Points 

### 1. Change of ep for BLT vs. PIT during their games in 2010 

```{r}
ep_blt_vs_pit <- full_with_elo %>% 
  filter(home_club_code %in% c("BLT","PIT"),
         visitor_club_code %in% c("BLT", "PIT"),
         season.x == "2010") %>%
  select(game_key, date, season.x, game_clock, possession_team, ep) 
```

```{r}
blt_vs_pit_2010 <- ggplot(data = ep_blt_vs_pit, aes(x = game_clock, y = ep, color = possession_team)) +
  geom_line() + 
  facet_wrap(facets = "date", nrow = 5) + 
  ggtitle(label = "Change of ep for BLT vs. PIT Games in 2010.", subtitle = "Shows Offensive Team Production") 
blt_vs_pit_2010
```


### Function that shows the Change of Ep
```{r}
# This function takes in team and year and outputs the expected points of the home team for each game of the in a given year.  

team_dep <- function(team, year) { 
  dep_data <- full_with_elo %>% 
    filter(home_club_code == team,
           season.x == year) %>% 
    select(game_key, date, season.x, game_clock, possession_team, ep)
  
  dep_plot <- ggplot(data = dep_data, aes(x = game_clock, y = ep, color = possession_team)) +
    geom_line() + 
    facet_wrap(facets = "date", nrow = 5) + 
    ggtitle(label = "Change of Expected Points in Games", subtitle = "Shows Offensive Team Production") 
  
  return(dep_plot) 
}
```

### 2. Change of Expected Points for the offensive team for all Baltimore games in 2010 
```{r}
# example using above team_dep(): 
team_dep("BLT", "2010")
```

## EDA: Using Elo Ratings 

### 1. Pittsburg's Elo Rating Before Games in 2019 (as home team)

```{r}
pit_full_home <- full_with_elo %>%
  filter(home_club_code == "PIT", 
         season.x == "2019") 
```

```{r}
pit_home_elo <- ggplot(data = pit_full_home, aes(x = as.numeric(week), y = elo1_pre)) +
  geom_point(aes(color = "Thursday"), size = 2, alpha = 0.6) +
  geom_line() + 
  theme_bw() +
  geom_text(aes(label=visitor_club_code), hjust=0, vjust=0) +
  ggtitle(label = "Pittsburg's Elo Rating Before Games in 2019", subtitle = "as the home team.") + 
  labs(x = "week", y="Winning Probability") 
pit_home_elo
```


### 2: Pittsburg's Elo Rating Before Games in 2019 (as home team)

```{r}
pit_full_visitor <- full_with_elo %>%
  filter(visitor_club_code == "PIT", 
         season.x == "2019") 
```

```{r}
pit_visitor_elo <- ggplot(data = pit_full_visitor, aes(x = as.numeric(week), y = elo2_pre)) +
  geom_line() + 
  geom_point(aes(color = is_thurs), size = 2, alpha = 0.6) +
  theme_bw() +
  geom_text(aes(label=home_club_code), hjust=0, vjust=0) +
  ggtitle(label = "Pittsburg's Elo Rating Before Games in 2019", subtitle = "as the visitor team.") + 
  labs(x = "week", y="Winning Probability")
pit_visitor_elo
```

### 3. Change in Elo of Home and Visitor Team in a Game. 

```{r}
ggplot(data = full_with_elo, aes(x = elo1_pre)) +
  geom_density(kernel = "rectangular") +
  ggtitle("Distribution of Home ELO")
```

#plot epa on thurs vs sunday - density
```{r}
ggplot(data =   full_with_elo, aes(x = elo2_pre)) +
  geom_density(kernel = "rectangular", aes(color = game_day)) +
  ggtitle("Distribution of Home ELO by gameday") 
```

```{r}
full_with_elo %>% 
  distinct(game_key, .keep_all = TRUE) %>% 
  arrange(home_club_code) %>% 
  filter(season.x == "2019")
```

```{r}
elo_afc <- full_with_elo %>% 
  select(game_date, is_thurs, home_club_code, visitor_club_code, elo1_pre, game_day) %>% 
  filter(game_day == "Thursday" | game_day == "Sunday") %>% 
  filter(home_club_code %in% c("BUF", "MIA", "NEP", "NYJ",
                             "BLT", "CIN", "CLE", "PIT",
                             "HOU", "IND", "JAC", "TEN",
                             "DEN", "KC", "OAK"),
         visitor_club_code %in% c("BUF", "MIA", "NEP", "NYJ",
                                "BLT", "CIN", "CLE", "PIT",
                                "HOU", "IND", "JAC", "TEN",
                                "DEN", "KC", "OAK")) %>% 
  group_by(home_club_code, is_thurs) %>% 
  distinct(game_date, elo1_pre) %>% 
  arrange(game_date) %>% 
  summarise(elo_rank = mean(elo1_pre)) 
```


```{r}
afc_elo_home <- ggplot(data = elo_afc, aes(x = home_club_code, y = elo_rank, fill = as.factor(is_thurs))) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle=90)) +
  ggtitle(label = "Comparing mean elo rankings", subtitle = "Offensive Production on Thursday  vs. Sunday") +
  labs(fill = "Is_Thurs")
afc_elo_home
```

```{r}
elo_afc_season <- full_with_elo %>% 
  select(game_date, is_thurs, home_club_code, visitor_club_code, elo1_pre, game_day, season.x) %>% 
  filter(game_day == "Thursday" | game_day == "Sunday") %>% 
  filter(home_club_code %in% c("BUF", "MIA", "NEP", "NYJ",
                             "BLT", "CIN", "CLE", "PIT",
                             "HOU", "IND", "JAC", "TEN",
                             "DEN", "KC", "OAK"),
         visitor_club_code %in% c("BUF", "MIA", "NEP", "NYJ",
                                "BLT", "CIN", "CLE", "PIT",
                                "HOU", "IND", "JAC", "TEN",
                                "DEN", "KC", "OAK")) %>% 
  group_by(home_club_code, season.x, is_thurs) %>% 
  summarise(elo_rank = mean(elo1_pre)) 
```

```{r}
elo_rank_season <- ggplot(data = elo_afc_season, aes(x = home_club_code, y = elo_rank, fill = as.factor(is_thurs))) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle=90)) +
  facet_wrap(vars(season.x), nrow = 3) +
  ggtitle(label = "Comparing mean elo winning probabilities", subtitle = "Winning Probabilities on Thursday  vs. Sunday") +
  labs(fill = "Is_Thurs")
elo_rank_season
```

```{r}
elo_win_afc <- full_with_elo %>% 
  select(game_date, is_thurs, home_club_code, visitor_club_code, elo1_pre, elo_prob1, game_day, season.x) %>% 
  filter(game_day == "Thursday" | game_day == "Sunday") %>% 
  filter(home_club_code %in% c("BUF", "MIA", "NEP", "NYJ",
                             "BLT", "CIN", "CLE", "PIT",
                             "HOU", "IND", "JAC", "TEN",
                             "DEN", "KC", "OAK"),
         visitor_club_code %in% c("BUF", "MIA", "NEP", "NYJ",
                                "BLT", "CIN", "CLE", "PIT",
                                "HOU", "IND", "JAC", "TEN",
                                "DEN", "KC", "OAK")) %>% 
  group_by(home_club_code, is_thurs, season.x) %>% 
  summarise(win_prob = mean(elo_prob1)) 
```

```{r}
elo_win_afc_plot <- ggplot(data = elo_win_afc, aes(x = home_club_code, y = win_prob, fill = as.factor(is_thurs))) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle=90)) +
  facet_wrap(vars(season.x), nrow = 3) +
  ggtitle(label = "Comparing mean elo winning probabilities", subtitle = "Winning Probabilities on Thursday  vs. Sunday") +
  labs(fill = "Is_Thurs")
elo_win_afc_plot
```


